A Framework for Discrete Change

### 2008-11-09 papers/individual/individual.tex
updated individual paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@237 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% problematic aspects of this analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials. A silent assumption that is made in these analyses is that
behavior has stabilized after 140 trials of learning; this could very
well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Single participant choices analyzed.

\includegraphics[width=7cm]{graphs/igtdata4.pdf}

\begin{itemize}
	\item N=200
	\item 4 choice data, displayed in blocks of 20 trials
	\item optimal strategy is choosing C or D
\end{itemize}

Results are the 4-state model (best by AIC/BIC), model predicted
probabilities are in the figure ???. States are characterized by
different types of behavior, shifting from B/D strategy to C/D
(optimal) strategy.

\includegraphics[width=7cm]{graphs/igtmodels4.pdf}

>>>>

% problematic aspects of standard analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials.  A silent assumption that is made in these analyses is
that behavior has stabilized after 140 trials of learning; this could
very well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Single participant IGT learning data are analyzed with the aim to
establish 1) at which trial no more learning occurs, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the 4 possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constaint that was imposed on the models' parameters was that
there were designated begin and end states.  This means that the
initial state probabilities were fixed to one for the first state and
to zero for the remaning states.  For the transition matrix this means
that the final state was an absorbing state: the probability of
transitioning out of that state is zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion (AIC; Akaike, 1973).

\nocite{Akaike1973} % check the year!!

We chose to analyze two participants that used different strategies at
the end of the task as analyzed in the manner proposed by Huizenga et
al\.  (2007).  The first participants' data were best described by a
4-state model.  This model's transition matrix is: 
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbind.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~ref{tab:igt4} below.

\begin{table}
	\caption{Estimates for the Iowa Gambling Task model}
	\label{tab:igt4}
	\begin{tabular}{lcccccccc} \hline
		state & p(A) & p(B) & p(C) & p(D) \\
		$S_1$ & 0    & 0.06 & 0.31 & 0.63 \\
		$S_2$ & 0.33 & 0.67 & 0    & 0 \\ 
		$S_3$ & 0    & 0    & 1    & 0 \\
		$S_4$ & 0    & 0.06 & 0.94 & 0 
	\end{tabular}
\end{table}

The states have clear interpretations. States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses. States 1 and 2 are dominated C and D and A and B responses 
respectively. To get a clearer intpretation of these states it is
necessary to consider when they are visited during the learning
process. The viterbi algorithm (REFERENCE) provides us with the
posterior state sequence, i.e., the sequence of states that the
participant is in at each trial. This sequence is depicted in
Figure~\ref{fig:post4}.

\begin{figure}
\begin{center}
	\scalebox{0.9}{\includegraphics*{graphs/post4.pdf}}
	\label{fig:post4}
	\caption{Posterior state sequence for the 4 state model.}
\end{center}
\end{figure}

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

\begin{figure}
\begin{center}
	\scalebox{1}{\includegraphics*{graphs/igt4.pdf}}
	\label{fig:igt4}
	\caption{Data and model predicted proportions of door choices in
	the IGT.}
\end{center}
\end{figure}

The initial preference for B and D choices confirms the theory
expressed in Huizenga et al. (2007) that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices reprensents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D reponses generate equal
profits in the long run.
